Histogram of Gradient Orientations of Signal Plots applied to P300 Detection

### 2017-12-15 article.tex
Subo los cambios de Juliana

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{EL OBJETIVO DE ESTE PAPER, EL HILO CONDUCTOR, ES DEMOSTRAR QUE ESTE METODO QUE ESTA BASADO EN LA ESTRUCTURA DE LA SENIAL SE PUEDE UTILIZAR TAMBIEN PARA IDENTIFICAR EVENTOS TRANSIENTES COMO EL P300, Y QUE CON ESO SE PUEDE IMPLEMENTAR UN SPELLER DE P300.  The analysis of Electroencephalographic (EEG) signals is of ulterior importance to elucidate patterns that could improve the implementation of Brain Computer Interfaces (BCI). These systems are meant to provide alternative pathways to transmit volitional information which could potentially enhance the quality of life of patients affected by neurodegenerative disorders or improve Human Computer Interaction systems.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) because they can be elicited by external stimuli and used to implement spellers, to control external devices or even avatars in virtual reality environments.  This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from the images of the plots of the signals.  It also aims to provide a framework to analyze, characterize and classify EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients and an own dataset for healthy subjects.}

@@

In~\citep{Ramele2016},  authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves~\citep{Ben-Simon2013}  and Motor Imagery rolandic central $\mu$ rhythms~\citep{Hartman2005} using the histogram of gradient orientations of signal plots. Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  Its validity is verified by processing offline data for ALS patients and for healthy subjects. 
The proposed approach is based on the morphological analysis of the shape of the EEG signal~\citep{Alvarado-Gonzalez2016,Yamaguchi2009} and it was built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots~\citep{Hartman2005}.

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of gradient orientations extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two public datasets.

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradient Orientations of the Signal Plot method is explained. Section~\ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.

\section{Materials and Methods}

The P300~\citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around $300$ ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm~\cite{WolpawJonathanR2012} and it is consistent across different subjects. It has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around $-15$ db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity~\citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix~\citep{Farwell1988}. Fig.~\ref{fig:p300matrix} shows an example of the Speller Matrix used in the OpenVibe Open Source software~\citep{Renard2010}, where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

@@

The underlying idea of this method is to generate templates for the signal \citep{Ramele2016}, based on the image plot.  Hence, the first step is the processing of the EEG signal.

\subsubsection{Processing Pipeline} \label{Pipeline}

\begin{itemize}
\item \textbf{Preprocessing}: The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The processing pipeline starts by applying a notch filter to the raw signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the $6$ rows plus the $6$ columns, a basic artefact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

@@

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot \frac{( x(t,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor, t\geq 0,c \in \{1,2,\dots,8\}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related with the image scale. On the other hand, $ x(t,c) $ is the point-to-point averaged EEG matrix for time $t$ and for channel $c$. Lastly, $\bar{x}(c)$ and $ \sigma (c) $ are the mean and standard deviation of $x$.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject. For each illuminated row or column, an image is obtained.
\item Repeat item 1 $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t$ and $c$ correspond to time and chanel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item Compute  
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c), t\geq 0, c \in \{1,\dots,8\}
\label{averaging}
\end{equation}  
for each row and column.

\item Plot the signal $x(t,c)$, according Section~\ref{Plot}. The images $I^{row}_1, \dots, I^{row}_6$ and $I^{col}_7,\dots,I^{col}_{12}$ are obtained for rows and columns, respectively.

@@

\begin{equation}
\hat{r} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(d^{row}_u)}^{} \left\lVert q -  d^{row}_u \right\rVert ^2
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\hat{c} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(d^{col}_u)}^{} \left\lVert q -  d^{col}_u \right\rVert ^2
\label{eq:multiclassificationcol}
\end{equation}

where $NN_T(d^l_u), \;l\in\{row,{col}\}$ is the set of the $k$ nearest neighbors to $d^l_u, \;l\in\{row,{col}\}$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l\in\{row,{col}\}$.  By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{r} $ and column $ \hat{c} $.

@@

\end{enumerate}

Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}

>>>>

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{%EL OBJETIVO DE ESTE PAPER, EL HILO CONDUCTOR, ES DEMOSTRAR QUE ESTE METODO QUE ESTA BASADO EN LA ESTRUCTURA DE LA SENIAL SE PUEDE UTILIZAR TAMBIEN PARA IDENTIFICAR EVENTOS TRANSIENTES COMO EL P300, Y QUE CON ESO SE PUEDE IMPLEMENTAR UN SPELLER DE P300.  
The analysis of Electroencephalographic (EEG) signals is of ulterior importance to elucidate patterns that could improve the implementation of Brain Computer Interfaces (BCI). These systems are meant to provide alternative pathways to transmit volitional information which could potentially enhance the quality of life of patients affected by neurodegenerative disorders or improve Human Computer Interaction systems.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) because they can be elicited by external stimuli and used to implement spellers, to control external devices or even avatars in virtual reality environments.  This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from the images of the plots of the signals.  It also aims to provide a framework to analyze, characterize and classify EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients and an own dataset for healthy subjects.}

@@

In~\citep{Alvarado-Gonzalez2016,Yamaguchi2009} an approach based on the morphological analysis of the shape of the EEG is presented. In this work the authors use slop chain code.

In the article~\citep{Ramele2016},  the authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves  and Motor Imagery rolandic central $\mu$ rhythms using the histogram of gradient orientations of signal plots.  Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  
The proposed approach is based on the morphological analysis of the shape of the EEG signal, but using histogram descriptors. The method is built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots, as it is described in~\citep{Hartman2005}.

This paper reports a method to, (1) classify P300 signals based on the identification of their structure in the shape domain using histograms of gradient orientations extracted from the image of signal plots, and (2) describe the way in which this classification can be used to implement an offline P300-based BCI Speller application using two public datasets. Its validity is verified by processing offline data for ALS patients and for healthy subjects.

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradient Orientations of the Signal Plot method is explained. Section~\ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.
%This kind of EEG events are described in~\citep{Hartman2005} and~\citep{Ben-Simon2013}.
\section{Materials and Methods}

The P300~\citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around $300$ ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm~\cite{WolpawJonathanR2012} and it is consistent across different subjects. It has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around $-15$ db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity~\citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix~\citep{Farwell1988}. Fig.~\ref{fig:p300matrix} shows an example of the Speller Matrix used in the OpenVibe Open Source software~\citep{Renard2010}, where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or a column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

@@

%The underlying idea of this method is to generate templates for the signal \citep{Ramele2016}, based on the image plot.  Hence, the first step is the processing of the EEG signal.
In this section, the signal preprocessing, the method for generating images from signal plots, the feature extraction procedure and the speller matrix identification are described.

\subsubsection{Preprocessing Pipeline} \label{Pipeline}

\begin{itemize}
\item \textbf{Signal Enhancement}: The preprocessing stage consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. The processing pipeline starts by applying a notch filter to the raw signal, a 
$4$th degree $10$ Hz lowpass Butterworth filter and finally a decimation with a Finite Impulse Response (FIR) filter of order $30$ from the original sampling frequency down to $16$ Hz \citep{Krusienski2006}.
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the $6$ rows plus the $6$ columns, a basic artefact elimination procedure is implemented by removing the entire sequence when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the $12$ stimuli,  a window of $1$ second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments should contain the P300 ERP signature time-locked to the flashing stimulus, one for the row, and one for the column.
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

@@

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot \frac{( x(t,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor, t \in \{0, 1, \dots, F_s\},c \in \{1,2,\dots,8\}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related with the image scale. On the other hand, $ x(t,c) $ is the point-to-point averaged EEG matrix for time $t$ and for channel $c$. The parameter $F_s$ is the sampling frequency. Lastly, $\bar{x}(c)$ and $ \sigma (c) $ are the mean and standard deviation of $x$.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject. For each illuminated row or column, an image is obtained.
\item Repeat item 1 $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t$ and $c$ correspond to time and chanel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute  
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c), t\geq 0, c \in \{1,\dots,8\}
\label{averaging}
\end{equation}  
for each row and column.

\item \label{paso4}Plot the signal $x(t,c)$, according Section~\ref{Plot}.

\item Repeat steps~\ref{paso3} and~\ref{paso4} in order to generate the images $I^{row}_1, \dots, I^{row}_6$ and $I^{col}_7,\dots,I^{col}_{12}$ for rows and columns, respectively.

@@

\begin{equation}
\hat{row} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(d^{row}_u)}^{} \left\lVert q -  d^{row}_u \right\rVert ^2
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\hat{col} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(d^{col}_u)}^{} \left\lVert q -  d^{col}_u \right\rVert ^2
\label{eq:multiclassificationcol}
\end{equation}

where $NN_T(d^l_u), \;l\in\{row,{col}\}$ is the set of the $k$ nearest neighbors to $d^l_u, \;l\in\{row,{col}\}$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l\in\{row,{col}\}$.

@@

\end{enumerate}
By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $. 
Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments $S_i$ are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}
